Storm Event Effects on Population and Economic Health
========================================================

## Synopsis
In this report, I aim to describe which storm events are the most damaging to population and economic health. My hypothesis is that tornadoes cause the most damages to both population and economic health. To investigate my hypothesis, I obtained data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including estimates of any fatalities, injuries, and property damages.From this data, we found that tornadoes are the most damaging to population health and flooding caused the most economic consequences.  

## Data Retrieval
Data retrieval and store raw CSV
```{r Download.Store.Storm.Data,cache=TRUE}
#Download File and Save Date Downloaded
NOAA.data.file <- download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",destfile="StormData.csv.bz2",method="curl")
dateDownloaded <- date()

#Unzip the bz2 archive
open.connection <- bzfile("StormData.csv.bz2","r")
NOAA.data <- read.csv(open.connection)
close(open.connection)
unlink("StormData.csv.bz2")
```
## Data Processing
For the start of my data processing, I want to only keep the columns that pertain to population health and economic damages. Therefore, I want to keep the following columns in my dataframe: "COUNTY","COUNTYNAME","STATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP","REMARKS".
```{r Data.Resizing,cache=TRUE}
#Subsetting the Data Frame to only include columns relevant for analysis
keeps <-c("COUNTY","COUNTYNAME","STATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP","REMARKS")
NOAA.data.keeps <-NOAA.data[,keeps]
```
Next, I want to keep only data that show fatalities, injuries, property damage, or crop damage.
```{r Data.Damages}
NOAA.data.with.damages <- subset(NOAA.data.keeps,
                                 NOAA.data.keeps$FATALITIES > 0 | NOAA.data.keeps$INJURIES > 0 | NOAA.data.keeps$PROPDMG > 0 | NOAA.data.keeps$CROPDMG >0) 
```
-General cleaning
```{r General.Cleaning.Data}
#Change all casing to upper
NOAA.data.with.damages$EVTYPE<- toupper( as.character(NOAA.data.with.damages$EVTYPE)) 
#Remove spaces from the front of EVTYPE
NOAA.data.with.damages$EVTYPE<-sub("^(\\s)+","",NOAA.data.with.damages$EVTYPE)
#Remove extra spaces from EVTYPE
NOAA.data.with.damages$EVTYPE<-gsub('[[:space:]]+',' ',NOAA.data.with.damages$EVTYPE)
```
-Renaming data so that naming style is the same
```{r Renaming.Data}
#Renaming various mispellings of THUNDERSTORM
NOAA.data.with.damages$EVTYPE<-sub("(TSTM)|(THUDERSTORM)|(TUNDERSTORM)|(THUNDERSTROM)|(THUNDERTORM)|(THUNERSTORM)|(THUNDEERSTORM)|(THUNDERESTORM)","THUNDERSTORM",NOAA.data.with.damages$EVTYPE)
#Renaming TORNDAO as TORNADO
NOAA.data.with.damages$EVTYPE<-sub("TORNDAO","TORNADO",NOAA.data.with.damages$EVTYPE)
#Renaming Avalance to Avalanche
NOAA.data.with.damages$EVTYPE<-sub("AVALANCE","AVALANCHE",NOAA.data.with.damages$EVTYPE)
#Renaming MIRCO to MICRO
NOAA.data.with.damages$EVTYPE<-sub("MIRCO","MICRO",NOAA.data.with.damages$EVTYPE)
#Renaming CSTL to COASTAL
NOAA.data.with.damages$EVTYPE<-sub("CSTL","COASTAL",NOAA.data.with.damages$EVTYPE)
#Renaming WINDCHILL to WIND CHILL
NOAA.data.with.damages$EVTYPE<-sub("WINDCHILL","WINDCHILL",NOAA.data.with.damages$EVTYPE)
#Renaming various misordering of FLASH FLOOD
NOAA.data.with.damages$EVTYPE<-sub("(FLOOD FLASH)|(FLOOD\\/FLASH)","FLASH FLOOD",NOAA.data.with.damages$EVTYPE)
#Renaming FLD to FLOOD
NOAA.data.with.damages$EVTYPE<-sub("FLD","FLOOD",NOAA.data.with.damages$EVTYPE)
#Renaming LIGHTING to LIGHTNING
NOAA.data.with.damages$EVTYPE<-sub("LIGHTING","LIGHTNING",NOAA.data.with.damages$EVTYPE)
#Renaming FLOES to FLOWS
NOAA.data.with.damages$EVTYPE<-sub("FLOES","FLOWS",NOAA.data.with.damages$EVTYPE)
#Renaming Non-Thunderstorm Wind to Wind 
NOAA.data.with.damages$EVTYPE<-sub("(NON-THUNDERSTORM)|(NON THUNDERSTORM)","",NOAA.data.with.damages$EVTYPE)
```
-Create column to group events into many of the categories listed in FAQ
```{r Group.Data,cache=TRUE}
#Create Subset for Tide or Storm Surge Storm Events
NOAA.data.tides <- NOAA.data.with.damages[grepl("(tide)|(surge)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.tides$General.Storm.Event <-"Storm Surge/Tide"

#Create Subset for Avalanche Storm Events
NOAA.data.avalanche <- NOAA.data.with.damages[grepl("avalanche",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.avalanche$General.Storm.Event <-"Avalanche"

#Create Subset for Blizzard Storm Events
NOAA.data.blizzard <- NOAA.data.with.damages[grepl("blizzard",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.blizzard$General.Storm.Event <-"Blizzard"

#Create Subset for Coastal Storm Events
NOAA.data.coastal <- NOAA.data.with.damages[grepl("(coastal)|(beach) ",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.coastal$General.Storm.Event <-"Coastal"

#Create Subset for Cold or Wind Chill Storm Events
NOAA.data.cold.windchill <- NOAA.data.with.damages[grepl("(cold)|(wind chill)|(hypothermia)|(low temperature)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.cold.windchill$General.Storm.Event <-"Cold/Wind Chill"

#Create Subset for Debris Flow Storm Events
NOAA.data.debrisflow <- NOAA.data.with.damages[grepl("(slide)|(flows)|(slump)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.debrisflow$General.Storm.Event <-"Debris Flow"

#Create Subset for Dense Fog Storm Events
NOAA.data.densefogsmoke <- NOAA.data.with.damages[grepl("(fog)|(smoke)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.densefogsmoke$General.Storm.Event <-"Dense Fog or Freezing Fog or Dense Smoke"

#Create Subset for Drought Storm Events
NOAA.data.drought <- NOAA.data.with.damages[grepl("(drought)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.drought$General.Storm.Event <-"Drought"

#Create Subset for Dust Storm Events
NOAA.data.duststorm <- NOAA.data.with.damages[grepl("(dust devil)|(dust storm)|(dust)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.duststorm$General.Storm.Event <-"Dust Devil or Dust Storm"

#Create Subset for Heat Storm Events
NOAA.data.heat <- NOAA.data.with.damages[grepl("(heat)|(hyperthermia)|(warm)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.heat$General.Storm.Event <-"Heat"

#Create Subset for Flood Storm Events
NOAA.data.flood <- NOAA.data.with.damages[grepl("(flood)|(high water)|(rapidly rising water)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.flood$General.Storm.Event <-"Flash Flood or Flood"

#Create Subset for Frost/Freeze Storm Events
NOAA.data.frost <- NOAA.data.with.damages[grepl("(frost)|(freeze)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.frost$General.Storm.Event <-"Frost/Freeze"

#Create Subset for Funnel Cloud Storm Events
NOAA.data.funnel <- NOAA.data.with.damages[grepl("(funnel)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.funnel$General.Storm.Event <-"Funnel Cloud"

#Create Subset for Hail Storm Events
NOAA.data.hail <- NOAA.data.with.damages[grepl("(hail)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.hail$General.Storm.Event <-"Hail or Marine Hail"

#Create Subset for Thunderstorm Storm Events
NOAA.data.thunderstorm <- NOAA.data.with.damages[grepl("(thunderstorm)|(thundersnow)|(gustnado)|(apache)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.thunderstorm$General.Storm.Event <-"Thunderstorm Winds or Marine Thunderstorm Winds"

#Create Subset for Tornado Storm Events
NOAA.data.tornado <- NOAA.data.with.damages[grepl("(tornado)|(landspout)|(waterspout)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.tornado$General.Storm.Event <-"Tornado"

#Create Subset for Wind Storm Events
NOAA.data.wind <- NOAA.data.with.damages[grepl("(wind)|(microburst)|(downburst)|(whirlwind)|(turbulence)|(marine mishap)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.wind <- rbind(NOAA.data.wind,{subset(NOAA.data.with.damages,NOAA.data.with.damages$EVTYPE=="HIGH")})
NOAA.data.wind$General.Storm.Event <-"Wind"

#Create Subset for Rain Storm Events
NOAA.data.rain <- NOAA.data.with.damages[grepl("(rain)|(shower)|(wet)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.rain$General.Storm.Event <-"Rain"

#Create Subset for Snow Storm Events
NOAA.data.snow <- NOAA.data.with.damages[grepl("(snow)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.snow$General.Storm.Event <-"Snow"

#Create Subset for Surf Storm Events
NOAA.data.surf <- NOAA.data.with.damages[grepl("(surf)|(swells)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.surf$General.Storm.Event <-"Surf"

#Create Subset for Hurricane(Typhoon) Storm Events
NOAA.data.hurricane <- NOAA.data.with.damages[grepl("(hurricane)|(typhoon)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.hurricane$General.Storm.Event <-"Hurricane(Typhoon)"

#Create Subset for Lightning Storm Events
NOAA.data.lightning <- NOAA.data.with.damages[grepl("(lightning)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.lightning$General.Storm.Event <-"Lightning"

#Create Subset for Ice Storm Events
NOAA.data.ice <- NOAA.data.with.damages[grepl("(ice)|(icy)|(glaze)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.ice$General.Storm.Event <-"Ice"

#Create Subset for Tropical Depression Storm Events
NOAA.data.tropicaldepression <- NOAA.data.with.damages[grepl("(tropical depression)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.tropicaldepression$General.Storm.Event <-"Tropical Depression"

#Create Subset for Tropical Storm Events
NOAA.data.tropicalstorm <- NOAA.data.with.damages[grepl("(tropical storm)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.tropicalstorm$General.Storm.Event <-"Tropical Storm"

#Create Subset for Winter Storm/Weather Storm Events
NOAA.data.winter <- NOAA.data.with.damages[grepl("(winter storm)|(winter weather)|(wintry))",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.winter$General.Storm.Event <-"Winter Storm/Weather"

#Create Subset for Different Fire Storm Events
NOAA.data.fire <- NOAA.data.with.damages[grepl("(fire)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.fire$General.Storm.Event <-"Fire"

#Create Subset for Seiche Storm Events
NOAA.data.seiche <- NOAA.data.with.damages[grepl("(seiche)|(wave)|(high seas)|(heavy seas)|(rough seas)|(marine accident)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.seiche$General.Storm.Event <-"Seiche"

#Create Subset for Volcanic Ash Storm Events
NOAA.data.volcanicash <- NOAA.data.with.damages[grepl("(volcanic)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.volcanicash$General.Storm.Event <-"Volcanic Ash"

#Create Subset for Mix Precipitation Storm Events
NOAA.data.mixprecip <- NOAA.data.with.damages[grepl("(freezing drizzle)|(freezing spray)|(mix)|(precipitation)|(sleet)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.mixprecip$General.Storm.Event <-"Mixed Precipitation"

#Create Subset for RIP Current Storm Events
NOAA.data.ripcurrent <- NOAA.data.with.damages[grepl("(rip current)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.ripcurrent$General.Storm.Event <-"RIP Currents"

#Create Subset for Tsunami Storm Events
NOAA.data.tsunami <- NOAA.data.with.damages[grepl("(tsunami)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.tsunami$General.Storm.Event <-"Tsunami"

#Create Subset for Other Storm Events
NOAA.data.other <- NOAA.data.with.damages[grepl("(other)|(dam)|(drown)",
                                                  NOAA.data.with.damages$EVTYPE,
                                                  ignore.case=TRUE)  , ]
NOAA.data.dust <- NOAA.data.other[grepl("dust",NOAA.data.other$REMARKS,ignore.case=TRUE),]
NOAA.data.dust$General.Storm.Event <- "Dust Devil or Dust Storm"
NOAA.data.rainother <- NOAA.data.other[grepl("rain",NOAA.data.other$REMARKS,ignore.case=TRUE),]
NOAA.data.rainother$General.Storm.Event <- "Rain"
NOAA.data.windother <- NOAA.data.other[grepl("wind",NOAA.data.other$REMARKS,ignore.case=TRUE),]
NOAA.data.windother$General.Storm.Event <- "Wind"
NOAA.data.other$General.Storm.Event <-"Other"

#Combine all the subsets into one overall dataframe
NOAA.data.categorized <-rbind(NOAA.data.tides,NOAA.data.avalanche,NOAA.data.blizzard,NOAA.data.coastal,NOAA.data.cold.windchill,NOAA.data.debrisflow,NOAA.data.densefogsmoke,NOAA.data.drought,NOAA.data.duststorm,NOAA.data.dust,NOAA.data.heat,NOAA.data.flood,NOAA.data.frost,NOAA.data.funnel,NOAA.data.hail,NOAA.data.thunderstorm,NOAA.data.tornado,NOAA.data.wind,NOAA.data.rain,NOAA.data.rainother,NOAA.data.snow,NOAA.data.surf,NOAA.data.hurricane,NOAA.data.lightning,NOAA.data.ice,NOAA.data.tropicaldepression,NOAA.data.tropicalstorm,NOAA.data.winter,NOAA.data.fire,NOAA.data.seiche,NOAA.data.volcanicash,NOAA.data.mixprecip,NOAA.data.ripcurrent,NOAA.data.tsunami,NOAA.data.windother)
```
 
## Results
### Population Health
```{r Health.Summary,cache=TRUE}
library(plyr)
NOAA.data.health.summary <- ddply(NOAA.data.categorized,
                              ~General.Storm.Event,
                              summarize,
                              totalFatalities=sum(FATALITIES,na.rm=TRUE),totalInjuries=sum(INJURIES,na.rm=TRUE))
NOAA.data.health.summary<-NOAA.data.health.summary[order(-NOAA.data.health.summary$totalFatalities,-NOAA.data.health.summary$totalInjuries),]

head(NOAA.data.health.summary)
```
-Plot top 10
```{r Top.Ten.Health}
library(ggplot2)
library(reshape2)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
print({
  standard.plot <- ggplot(melt(NOAA.data.health.summary[1:10,],id.vars="General.Storm.Event",measure.vars=c("totalInjuries","totalFatalities")),aes(General.Storm.Event,value,fill=variable))
  plot.level.one <- standard.plot+geom_bar(stat="identity")+scale_fill_manual(values=cbPalette)
  plot.level.two <- plot.level.one+labs(title="Population Health Effects of Storm Events")
  plot.level.three <- plot.level.two + labs(x="Top Ten General Storm Events",y="Total Occurences")
  plot.level.three+theme(axis.text.x=element_text(angle=90,hjust=1))
  })
```
### Economic Consequences
-Add columns for value of Damage Exp
```{r Convert.Damage.Exp,cache=TRUE}
#Add column for Property Damage True Cost and True Exponent Value
levels(NOAA.data.with.damages$PROPDMGEXP)
map_column<-rep(c(1,1,1,1,1e0,1e1,1e2,1e3,1e4,1e5,1e6,1e7,1e8,1e9,1e2,1e2,1e3,1e6,1e6))
NOAA.data.categorized$PROPDMGEXPVAL<-map_column[NOAA.data.categorized$PROPDMGEXP]

#Add column for Crop Damage True Cost and True Exponent Value
levels(NOAA.data.with.damages$CROPDMGEXP)
map_column<-rep(c(1,1,1e0,1e2,1e9,1e3,1e3,1e6,1e6))
NOAA.data.categorized$CROPDMGEXPVAL<-map_column[NOAA.data.categorized$CROPDMGEXP]
```
-Add columns for Overall Costs
```{r Overall.Dmg.Cost,cache=TRUE}
#Add column for Property Damage Cost
NOAA.data.categorized$PROPDMGCOST<-NOAA.data.categorized$PROPDMG*NOAA.data.categorized$PROPDMGEXPVAL

#Add column for Crop Damage Cost
NOAA.data.categorized$CROPDMGCOST<-NOAA.data.categorized$CROPDMG*NOAA.data.categorized$CROPDMGEXPVAL

#Add column for Overall Damages Cost
NOAA.data.categorized$OVERALLDMGCOST<-NOAA.data.categorized$PROPDMGCOST+NOAA.data.categorized$CROPDMGCOST
```

-Summarize Data
```{r Economic.Summary,cache=TRUE}
library(plyr)
NOAA.data.economic.summary <- ddply(NOAA.data.categorized,
                              ~General.Storm.Event,
                              summarize,
                              totalOverallDmg=sum(OVERALLDMGCOST,na.rm=TRUE))
NOAA.data.economic.summary<-NOAA.data.economic.summary[order(-NOAA.data.economic.summary$totalOverallDmg),]

head(NOAA.data.economic.summary)
```
-Plot top 10
```{r Top.Ten.Economic}
library(ggplot2)
print({
  standard.plot <- ggplot(NOAA.data.economic.summary[1:10,],aes(General.Storm.Event,totalOverallDmg))
  plot.level.one <- standard.plot+geom_bar(stat="identity")
  plot.level.two <- plot.level.one+labs(title="Economic Health Effects of Storm Events")
  plot.level.three <- plot.level.two + labs(x="Top Ten General Storm Events",y="Total Cost of Damages")
  plot.level.three+theme(axis.text.x=element_text(angle=90,hjust=1))
  })
```
